---
title: "Approximating the Gradient using First Differences"
author: "Gary Mitchell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(kableExtra)
require(reticulate)
suppressMessages(require(numform))
```

In a previous tutorial we learned to use gradient descent to find a minimum or maximum of a function. If the function is strictly convex or strictly concave, gradient descent will find the globally optimal solution. If it is nont strictly convex/concave, it may find a local optimum, or even diverge.  

Gradient descent, as we learned it, depends on being able to find the analytical form for the derivative. However, for some functions it is either computationally very expensive or not possible to find an analytical form for the derivative.

This is where first differences are useful.

## First Differences

Recall that the definition of the derivative is:

$$\frac{d f(x)}{dx} = \lim_{\Delta x \to 0} \frac{f(x+\Delta x)-f(x)}{\Delta x}$$

This definition directly yields the idea of using first differences to approximate the derivative; the numerator is the difference of the function evaluated at two points. If we let $\Delta x = 1$, then we have:

$$\frac{df(x)}{dx}\approx \Delta f(x) = f(x+\Delta x) - f(x)$$
That's it! That's the whole idea behind using first differences to approximate the gradient. Using the first difference, we calculate a step as follows:

$$x_{n+1} = f(x_n) - \eta \Delta f(x_n) = x_n - \eta\bigl[f(x + \Delta x) - f(x)\bigr]$$

*Technically*, we could use any $\Delta x$ as long as we divided the difference by $\Delta x$. However, for small $\Delta x$, the calculation can become unstable, so we generally settle for $\Delta x \geq 1$.


## Example: Finding a Global Minimum

To use first differences in gradient descent, we simply use the first difference in place of the gradient.

```{r, echo=FALSE}
curve(x^2 +2*x - 5, -5, 3, col='purple',
      main=expression(x^2 + 2*x -5),
      ylab=expression(f(x)))
abline(h=0)

```

Here's our algorithm from the gradient descent tutorial. Note that *no changes* are required to the gradient descent code (from the previous tutorial) to use first differences to approximate the gradient.

```{python}
import numpy as np
import pandas as pd
import math

def gradient_descent(f, grad, initial_guess, eta=0.01, iter_limit=500, epsilon=0.00001):
  
  x = np.full(iter_limit + 1, np.nan)
  fx = np.full(iter_limit + 1, np.nan)
  fpx = np.full(iter_limit + 1, np.nan)
  step = np.full(iter_limit + 1, np.nan)
  x[0] = initial_guess
  
  for i in range(0, len(x) - 1):
    fx[i] = f(x[i])
    fpx[i] = grad(x[i])
    step[i] = eta * fpx[i]
    x[i+1] = x[i] - step[i]
    
    if math.isclose(f(x[i+1]), 0, abs_tol=epsilon) or \
       math.isclose(x[i+1] - x[i], 0, abs_tol=epsilon):
      n = np.arange(i+2)
      df = pd.DataFrame({'n': n,
                         '$x_n$': x[:i+2],
                         '$f(x)$': fx[:i+2],
                         "$f'(x)$": fpx[:i+2],
                         'Step Size': step[:i+2]})
      return(df)
  
  return math.nan

```

As we did before, we'll start at $x_0$ = 6, just to give us room to observe a few iterations. Note that we simply defined the gradient function to calculate the first difference (instead of an actual gradient)! No other changes are required.

```{python}

initial_guess = 6


def f(x):
  return x**2 + 2 * x - 5

def grad(x, dx=1):
  return (f(x+dx) - f(x)) / dx

df = gradient_descent(f, grad, initial_guess)
  

```

Finding the optimal solution to 4 decimal places required `r dim(py$df)[1]-1` iterations.

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 1.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

Let's try increasing the learning rate.

```{python}

initial_guess = 6

df = gradient_descent(f, grad, initial_guess, eta=0.1)
  

```

That's quite a bit better, with convergence happening in `r dim(py$df)[1]-1` iterations.


```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 2.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

## Another Example

Let's apply gradient descent using first differences to finding the optimal order quantity for an EOQ problem (i.e., minimize the total relevant cost, $TRC(Q)$):

$$TRC(Q) = \frac{Q v r}{2} + \frac{D A}{Q}$$

where: 
  $D =$ annual demand
  $A =$ aggregate order placement cost
  $v =$ unit acquisition cost
  $r =$ unit holding cost percentage rate.
  
The figure below illustrates $TRC(Q)$:

```{r, echo=FALSE}
D<-10000
A<-500
v<-12.95
r<-0.35

curve(x * v * r / 2 + D * A / x, 1000, 3000,
      ylab=expression(TRC(Q)),
      xlab=expression(Q),
      col='forestgreen')

```

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = D / 52


def f(Q):
  return Q * v * r / 2 + D * A / Q

def grad(Q, dQ=1):
  return f(Q + dQ) - f(Q)

df = gradient_descent(f, grad, initial_guess, eta=0.1, iter_limit=100000)
  

```

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 4.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

OK, so we converged in `r format(dim(py$df)[1] - 1, big.mark=",")` iterations with an initial guess of $Q_0 = `r py['initial_guess']`$, and a learning rate of $0.10$; with $\eta=0.01$ the algorithm failed to converge in 100,000 iterations! But `r format(dim(py$df)[1] - 1, big.mark=",")` iterations' hardly seems efficient. What happens if we change the initial guess to 1,000?

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = 1000

df = gradient_descent(f, grad, initial_guess, eta=0.1, iter_limit=100000)
  

```


```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 5.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

It *still* requires nearly 30,000 iterations to converge. Let's try a larger learning rate:

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = 1000

df = gradient_descent(f, grad, initial_guess, eta=1.0, iter_limit=100000)
  

```

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 6.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

It's getting better. One more time with $\eta=5.0$:

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = 1000

df = gradient_descent(f, grad, initial_guess, eta=5.0, iter_limit=100000)
  

```

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 7.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

OK, that's more like it. We converged in only `r format(dim(py$df)[1] - 1, big.mark=",")` iterations.

Just out of curiosity, since we know the exact optimal form for the EOQ model, let's compare:

$$Q^*_{exact} = `r round(sqrt(2 * D * A / (v * r)),3)`$$
```{python, echo=FALSE}

qstar = df.iloc[-1]['$x_n$']

```


$$Q^*_{GD} = `r round(py['qstar'],3)`$$