---
title: "Constructing Confidence Intervals for Simulation Results"
author: "Gary Mitchell"
date: "2026-09-08"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(reticulate)
require(kableExtra)

```

The appropriate method for calculating confidence intervals for simulated results depends on the nature of the system being simulated. Two factors impact the process for constructing confidence intervals:

- Whether or not you need to allow for a *warm-up* period before your simulated system achieves *steady state* operation
- Whether your simulated system operates continuously or on a cycle.

## The warm-up period

For example, a feasible queuing system must achieve a steady state at some point. However, before it achieves a steady state, there is a *warm-up period* during which the system's state is not representative. For example, when a Starbucks opens, the queue is empty (i.e., nobody in the system). It takes some amount of being open before the queuing system has achieved its steady state.

To address this, we typically *throw away* some observations at the beginning of a simulation; we need to discard enough observations so that the remaining observations reflect steady state for the system, but we don't want to throw away so many that we are incurring lots of calculation overhead for nothing. There is no exact number of *warm-up* observations to discard that applies to every simulation. In each case, you must to determine how many observations to discard. If the simulation runs quickly, then you have the luxury of simply simulating a lot of trials so that you can absorb the loss of a significant number of warm-up trials. If the simulation runs slowly, then you'll want to choose the number of discards more carefully, to avoid unnecessarily long run times.

*For both continuously operating and cycle systems, you must discard the warm-up trials before constructing confidence intervals.*

## Continuously operating systems

Queuing systems (at least queuing system models) are often presented based on the implicit assumption that the system is continuously operating. Some examples of valid continuously operating systems:

- A 24-hour grocery store
- Amazon.com's web site
- A hospital emergency room.

All three of these examples operate continuously. In general, continuously operating systems require warm-up periods (and therefore discarding warm-up trials).

## Cycle systems

A cycle system is one where a defined *cycle* is repeated on a regular basis. For example, a grocery store that opens from 6am until 11pm daily, operates on a daily cycle. In cycle systems, the system starts each cycle with a defined state. For example, the grocery store opening at 6am starts every day with 0 customers in the system.

Cycle systems may never achieve true *steady state* operation. For example, the length/duration of the cycle may be too short for the system to achieve steady state. In general, cycle system simulations generally do not require discarding warm-up trials. In fact, discarding warm-up trials for a cycle system simulation may lead to misleading results.

## Constructing confidence intervals

So, the first step in constructucting the confidence interval for a simulation is to discard warm-up trials, if necessary.

Next, we have to determine whether or not we will need to use the *batch means* method. For cycle systems, our simulation almost always involves simulating many *cycles* for the system, and constructing the confidence intervals is straight forward. 

For example, in project management, we would simulate the schedule for a project many times. Simulating the project from beginning to end would constitute a single cycle. We would simulate many cycles for the project (e.g., 10,000 trials) and the resulting simulated cycles would comprise our simulation results. In this case, each cycle is independent of the others, simplifying our work to construct a confidence interval. 

Given some metric $m$ for which we want to construct the confidence interval, where $m_i$ is the value of the metric for the $i^{th}$ trial or cycle, we calculate a confidence interval as follows:

- Find the mean and standard error for the metric
- Find the critical value for the confidence interval
- Calculate the confidence interval using the mean, standard error, and critical value

Finding the mean and standard error is straightforward:

$$\bar{m} = \frac{1}{n} \sum_{i=1}^n m_i$$

$$s_d = \sqrt{\frac{1}{n} \sum_{i=1}^n (m_i - \bar{m})^2}$$

Then the standard error is constructed using:

$$s_e = \frac{s_d}{\sqrt{n}}$$

where $n$ is the number of cycles/trials simulated.

Finding the critical value depends on $n$:

- For $n < 30$, use Student's t distribution
- for $n \geq 30$, use the normal distribution

For cycle system simulations, $n >> 30$ is almost always true, so we generally use the normal distribution. To find the $\alpha$ (e.g., 95%) confidence interval, we would use the following:

$$k = F^{-1}\Bigg(1 - \bigg(\frac{1-\alpha}{2}\bigg)\Bigg)$$

then the confidence interval is:

$$\bar{m} - k\cdot s_e \leq \mu \leq \bar{m} + k\cdot s_e$$

For $n < 30$, we would find the critical value as follows:

$$k = F^{-1}\Bigg(1 - \bigg(\frac{1-\alpha}{2}\bigg), \ n-1\Bigg)$$

where $F^{-1}(p, d)$ is the inverse cumulative probability for Student's t distribution with $d$ degrees of freedom.

## Using batch means for continuously operating systems

The approach to calculating the confidence interval shown above can be a bit more problematic with continuously operating systems. Let's say we wanted to simulate traffic at Amazon.com. To use the approach described above, we would have to:

- Run multiple *cycles* where each cycle reflects the queue's *steady state* operation over some time interval
- For each cycle, discard the *warm-up* trials
- Calculate the desired metric for each cycle
- Construct the confidence interval for the metric as shown above.

In this case, we'd be simulating and discarding a substantial number of trials. For each *cycle*, we'd have to discard enough trials to be confident that the remaining trial data represented steady state operation. leading to a large number of discarded trials. 

The alternative is to use the batch means approach:

1. Generate n trials for a single steady state simulation
1. Discard the $w$ warm-up trials (the remaining trials represent steady state operation)
1. Divide the remaining trials into $k$ batches of size $m=\frac{n-w}{k}$
1. For each batch $i$, calculate the *batch mean* $\bar{x}_i$ of the metric $x$
1. Using the batch means, calculate the *mean of the batch means* $\bar{\bar{x}}$ and the standard deviation of the means, $\sigma_{\bar{x}}$
1. Construct the confidence interval using $\bar{\bar{x}}$ and $\sigma_{\bar{x}}$

### Calculate the batch means

$$\bar{x}_i = \frac{1}{k} \sum_{j=1}^{m} x_j$$

### Calculate the means of the batch means


$$\bar{\bar{x}} = \frac{1}{k} \sum_{i=1}^{k} \bar{x}_i$$

### Calculate the batch standard deviations

$$\sigma_{\bar{x}}=\sqrt{\frac{1}{k} \sum_{i=1}^{k} (\bar{x}_i-\bar{\bar{x}})^2}$$

### Construct the confidence interval

Now, we use the mean and standard deviation of the batch, $\bar{\bar{x}}$ and $\sigma_{\bar{x}}$ to calculate the confidence interval:

$$\bar{\bar{x}} - k \cdot \sigma_{\bar{x}}\  \leq \ \bar{\bar{x}} \ \leq \ \bar{\bar{x}} + k \cdot \sigma_{\bar{x}}$$

where $k$ is the critical value chosen as described above.

### Python sample code

The Python code below defines a function for analyzing trial data presented in a DataFrame.

```{python echo=FALSE}

import numpy as np
import pandas as pd
import time
import math

from scipy.stats import norm
from scipy.stats import poisson
from scipy.stats import expon
import scipy.stats as stats

def sim(arr_dist, svc_time_dist, trials: int=10000, random_state=None):
    if random_state is None:
        # use system lock if no specific seed specified
        random_state = time.time()

    # create a random number generator instance and set random seed
    rng = np.random.default_rng(random_state)

    # initial next arrival and service completion clocks
    next_arr = arr_dist.rvs(1, rng)[0]
    next_svc_comp = math.inf

    # customers in system, sim time (i.e., time since start of sim), and StateTime (i.e., time in state)
    n = np.zeros(trials + 1)
    sim_time = np.zeros(trials + 1)
    state_time = np.zeros(trials + 1)

    # initialize debugging vectors to capture next arrival and svc completion times at beginning of
    # event
    arr_time_beg = np.zeros(trials + 1)
    arr_time_end = np.zeros(trials + 1)
    svc_time_beg = np.zeros(trials + 1)
    svc_time_end = np.zeros(trials + 1)

    # initialize the starting state
    sim_time[0] = 0.0
    n[0] = 0

    for i in range(1, trials + 1):
        # save debug information
        arr_time_beg[i] = next_arr
        svc_time_beg[i] = next_svc_comp

        if next_arr < next_svc_comp:
            # The next arrival is scheduled BEFORE the next service completion

            # increase the customers in the system
            n[i] = n[i-1] + 1

            # update the simulation clock (i.e., running time)
            sim_time[i] = next_arr

            # calculate the time for the next arrival
            next_arr = sim_time[i] + arr_dist.rvs(1, rng)[0]

            if math.isinf(next_svc_comp):
                # The system was empty, so we need to calculate the next system time.
                # inf is used to ensure that the next event must be an arrival if the
                # system is empty
                next_svc_comp = sim_time[i] + svc_time_dist.rvs(1, rng)[0]
        else:
            # The next service completion is scheduled before the next arrival
            # reduce the number of customers in the system
            n[i] = n[i-1] - 1

            # update the system clock to reflect the current sim time
            sim_time[i] = next_svc_comp

            # Update the next svc completion time based on the state.
            if n[i] > 0:
                # The queue is not empty, so calculate the next service completion starting
                # from the current sim time
                next_svc_comp = sim_time[i] + svc_time_dist.rvs(1, rng)[0]
            else:
                # The queue is empty, so ensure that the next event is an arrival by setting
                # next_svc_comp to infinity.
                next_svc_comp = math.inf

        # calculate time spent in state
        state_time[i-1] = sim_time[i] - sim_time[i-1]
        arr_time_end[i] = next_arr
        svc_time_end[i] = next_svc_comp

    # calculate final StateTime
    if next_arr < next_svc_comp:
        # the next event would be an arrival, so use next arrival time to determine the time spent
        # in the state following the last event
        state_time[i] = next_arr - sim_time[i]
    else:
        # the next event would be a service completion, so use next svc completion time to determine
        # the time spent in the state following the last event
        state_time[i] = next_svc_comp - sim_time[i]

    # construct and return results data
    df = pd.DataFrame({'Time': sim_time, 
                       'ArrTimeBeg': arr_time_beg, 
                       'SvcTimeBeg': svc_time_beg, 
                       'n': n, 
                       'StateTime': state_time,
                       'ArrTimeEnd': arr_time_end,
                       'SvcTimeEnd': svc_time_end})

    return df
  
```

```{python, echo=FALSE}

arr_dist = expon(scale=1/20)
svc_time_dist = expon(scale=1/25)
k = 31

rslts = sim(arr_dist, svc_time_dist, trials=20000 + k * 10000, random_state=100)

```


```{python }
def analyze_sim(rslts: pd.DataFrame, warmup_n: int=0, k: int=30, alpha=0.95) -> dict:

    # drop warmup period observations - drop the initial state + the specified warmup_n states
    steady_state = rslts.iloc[(warmup_n + 1):, ]

    # reshape trial data so that each row represents one batch
    n = steady_state['n'].to_numpy().reshape(k,-1)
    t = steady_state['StateTime'].to_numpy().reshape(k,-1)

    # calculate the number of customers waiting from the number of customers in system
    # this is already batched. This is the basis for Lq
    n_wait = (n - 1).clip(min=0)

    # now calculate the weighted average mean of each batch
    mean_n_i = np.sum(n * t, axis=1) / t.sum(axis=1)
    mean_n_wait_i = np.sum(n_wait * t, axis=1) / t.sum(axis=1)

    # now we have the batch means, let's calculate means and standard deviations of the means
    mean_n = mean_n_i.mean()
    sd_n = mean_n_i.std()
    sem_n = sd_n / math.sqrt(k)

    mean_n_wait = mean_n_wait_i.mean()
    sd_n_wait = mean_n_wait_i.std()
    sem_n_wait = sd_n_wait / math.sqrt(k)

    #find the critical value
    dof = k - 1
    if k > 30:
        # use student t distribution for small sample size
        k = norm.ppf(1 - (1 - alpha) / 2)
    else:
        # use normal distribution for larger sample sizes
        k = stats.t.ppf(1 - (1 - alpha) / 2, df=k)

    # calculate batch and overall metrics for L and Lq
    analysis = {'batch': 
                    {'E[L]': mean_n,
                     f'L {alpha:.1%} confid': (mean_n - k * sem_n, mean_n + k * sem_n),
                     'E[Lq]': mean_n_wait,
                     f'Lq {alpha:.1%} confid': (mean_n_wait - k * sem_n, mean_n_wait + k * sem_n)
                },
                'Overall': {'E[L]': steady_state['n'].dot(steady_state['StateTime']
                                                          / steady_state['StateTime'].sum()),
                            'E[Lq]': (steady_state['n'] - 1).clip(lower=0).dot(steady_state['StateTime']
                                                          / steady_state['StateTime'].sum())
                            }
                }

    return analysis

```


Here's a sample of the DataFrame from a $10,000$ trial simulation. When $n = 0$, i.e., the queue is empty, the nexet service time is set to $\infty$ to ensure that an arrival is the next event.

```{r, echo=FALSE}

r_rslts <- reticulate::py_to_r(py$rslts)
r_rslts[r_rslts == Inf] <- '$\\infty$'


#Render the table

kbl(r_rslts[1:10,],
    format = "html",
    booktabs=TRUE,
    escape = FALSE,
    caption = "Sample Trials") %>%
  kable_styling(bootstrap_options="striped",
                full_width=FALSE)


# head(r_rslts)

```

Now, let's run the analysis:

```{python}

analysis = analyze_sim(rslts, warmup_n=20000, k=k)
for k,v in analysis.items():
  print(k)
  
  for a,b in v.items():
    print(f'{a}: {b}')
    
    
```

Here are the *pretty* confidence intervals.

```{python, echo=FALSE}

batch = analysis['batch']
overall = analysis['Overall']

el = batch['E[L]']
elq = batch['E[Lq]']
ellb = batch['L 95.0% confid'][0]
elub = batch['L 95.0% confid'][1]
elqlb = batch['Lq 95.0% confid'][0]
elqub = batch['Lq 95.0% confid'][1]

```

$$E[L]: `r round(py$ellb, 2)` \leq `r round(py$el, 2)` \leq `r round(py$elub, 2)`$$

$$E[L_q]: `r round(py$elqlb, 2)` \leq `r round(py$elq, 2)` \leq `r round(py$elqub, 2)`$$

Note that the theoretically exact values are:

$$E[L] = 4.0$$
$$E[L_q] = 3.2$$

## Generating the simulated trials

While this tutorial is about constructing confidence intervals, you might be interested in the code that generates the trials:

```{python, eval=FALSE}
import numpy as np
import pandas as pd
import time
import math

from scipy.stats import norm
from scipy.stats import poisson
from scipy.stats import expon
import scipy.stats as stats

def sim(arr_dist, svc_time_dist, trials: int=10000, random_state=None):
    if random_state is None:
        # use system lock if no specific seed specified
        random_state = time.time()

    # create a random number generator instance and set random seed
    rng = np.random.default_rng(random_state)

    # initial next arrival and service completion clocks
    next_arr = arr_dist.rvs(1, rng)[0]
    next_svc_comp = math.inf

    # customers in system, sim time (i.e., time since start of sim), and state time (i.e., time in state)
    n = np.zeros(trials + 1)
    sim_time = np.zeros(trials + 1)
    state_time = np.zeros(trials + 1)

    # initialize debugging vectors to capture next arrival and svc completion times at beginning of
    # event
    arr_time_beg = np.zeros(trials + 1)
    arr_time_end = np.zeros(trials + 1)
    svc_time_beg = np.zeros(trials + 1)
    svc_time_end = np.zeros(trials + 1)

    # initialize the starting state
    sim_time[0] = 0.0
    n[0] = 0

    for i in range(1, trials + 1):
        # save debug information
        arr_time_beg[i] = next_arr
        svc_time_beg[i] = next_svc_comp

        if next_arr < next_svc_comp:
            # The next arrival is scheduled BEFORE the next service completion

            # increase the customers in the system
            n[i] = n[i-1] + 1

            # update the simulation clock (i.e., running time)
            sim_time[i] = next_arr

            # calculate the time for the next arrival
            next_arr = sim_time[i] + arr_dist.rvs(1, rng)[0]

            if math.isinf(next_svc_comp):
                # The system was empty, so we need to calculate the next system time.
                # inf is used to ensure that the next event must be an arrival if the
                # system is empty
                next_svc_comp = sim_time[i] + svc_time_dist.rvs(1, rng)[0]
        else:
            # The next service completion is scheduled before the next arrival
            # reduce the number of customers in the system
            n[i] = n[i-1] - 1

            # update the system clock to reflect the current sim time
            sim_time[i] = next_svc_comp

            # Update the next svc completion time based on the state.
            if n[i] > 0:
                # The queue is not empty, so calculate the next service completion starting
                # from the current sim time
                next_svc_comp = sim_time[i] + svc_time_dist.rvs(1, rng)[0]
            else:
                # The queue is empty, so ensure that the next event is an arrival by setting
                # next_svc_comp to infinity.
                next_svc_comp = math.inf

        # calculate time spent in state
        state_time[i-1] = sim_time[i] - sim_time[i-1]
        arr_time_end[i] = next_arr
        svc_time_end[i] = next_svc_comp

    # calculate final state time
    if next_arr < next_svc_comp:
        # the next event would be an arrival, so use next arrival time to determine the time spent
        # in the state following the last event
        state_time[i] = next_arr - sim_time[i]
    else:
        # the next event would be a service completion, so use next svc completion time to determine
        # the time spent in the state following the last event
        state_time[i] = next_svc_comp - sim_time[i]

    # construct and return results data
    df = pd.DataFrame({'Time': sim_time,
                       'ArrTimeBeg': arr_time_beg,
                       'SvcTimeBeg': svc_time_beg,
                       'n': n,
                       'StateTime': state_time,
                       'ArrTimeEnd': arr_time_end,
                       'SvcTimeEnd': svc_time_end})

    return df


```


 