---
title: "Gradient Descent"
author: "Gary Mitchell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(kableExtra)
require(reticulate)
suppressMessages(require(numform))
```

Gradient Descent is a numerical method for a minimum or maximum of a function. If the function is strictly convex or strictly concave, gradient descent will find the globally optimal solution. If it is nont strictly convex/concave, it may find a local optimum, or even diverge.  
The method is pretty straightforward. We proceed iteratively from one step to the next, starting with a current guess $x_n$:

$$x_{n+1} = x_n - \eta  f'(x_n)$$
where $\eta$ is the learning rate and $f'(x_n)$ is the first derivative of $f(x)$, the function we are optimizing, evaluated at $x_n$.

In gradient descent, the choice of the learning rate is crucial:

- Too large, and the algorithm is likely to diverge
- Too small, and it will take *forever* (i.e., many iterations) to converge

## Example: Finding a Global Minimum

For example, we can use gradient descent to find the minimum of the strictly convex function shown below.

```{r, echo=FALSE}
curve(x^2 +2*x - 5, -5, 3, col='purple',
      main=expression(x^2 + 2*x -5),
      ylab=expression(f(x)))
abline(h=0)

```

First, we find $f'(x)$:

$$f'(x) = \frac{df(x)}{dx} = \frac{d}{dx}(x^2 + 2x - 5) = 2x + 2$$

Let's start at $x_0$ = 6, just to give us room to observe a few iterations:

```{python}
import numpy as np
import pandas as pd
import math

def gradient_descent(f, grad, initial_guess, eta=0.01, iter_limit=500, epsilon=0.00001):
  
  x = np.full(iter_limit + 1, np.nan)
  fx = np.full(iter_limit + 1, np.nan)
  fpx = np.full(iter_limit + 1, np.nan)
  step = np.full(iter_limit + 1, np.nan)
  x[0] = initial_guess
  
  for i in range(0, len(x) - 1):
    fx[i] = f(x[i])
    fpx[i] = grad(x[i])
    step[i] = eta * fpx[i]
    x[i+1] = x[i] - step[i]
    
    if math.isclose(f(x[i+1]), 0, abs_tol=epsilon) or \
       math.isclose(x[i+1] - x[i], 0, abs_tol=epsilon):
      n = np.arange(i+2)
      df = pd.DataFrame({'n': n,
                         '$x_n$': x[:i+2],
                         '$f(x)$': fx[:i+2],
                         "$f'(x)$": fpx[:i+2],
                         'Step Size': step[:i+2]})
      return(df)
  
  return math.nan

```

In the *gradient_descent* function, we use a for loop to ensure algorithm termination; if the loop terminates without converging to a solution, the function returns *math.nan*. The code leaves out some usual error checking (i.e., checking for a zero derivative to avoid zero divide errors) in the interest of emphasizing the *functional* code for the algorithm.

We'll use 6 as our initial guess.

```{python}

initial_guess = 6


def f(x):
  return x**2 + 2 * x - 5

def grad(x):
  return 2 * x + 2

df = gradient_descent(f, grad, initial_guess)
  

```

As it turns out, finding the optimal solution to 4 decimal places required `r dim(py$df)[1]-1` iterations when Newton's Method found it in 1 iteration.

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 1.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

Let's try increasing the learning rate.

```{python}

initial_guess = 6

df = gradient_descent(f, grad, initial_guess, eta=0.1)
  

```

That's quite a bit better, with convergence happening in `r dim(py$df)[1]-1` iterations.


```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 2.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

What happens if we increase it even further?


```{python}

initial_guess = -6

df = gradient_descent(f, grad, initial_guess, eta=1.0)
  

```

Looking at the table below, we see that the algorithm fails to converge when the learning rate was too high.

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 3.'

py$df %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

# The Problem with Gradient Descent

As you can see, gradient descent is entirely dependent on:

- the initial guess
- the learning rate.

Finding the right combination of these two crucial parameters can be a challenge. For this reason, simple gradient descent (as we've discussed here) is rarely used in practice anymore.

## Another Example

Let's apply gradient descent to finding the optimal order quantity for an EOQ problem (i.e., minimize the total relevant cost, $TRC(Q)$):

$$TRC(Q) = \frac{Q v r}{2} + \frac{D A}{Q}$$

where: 
  $D =$ annual demand
  $A =$ aggregate order placement cost
  $v =$ unit acquisition cost
  $r =$ unit holding cost percentage rate.
  
The figure below illustrates $TRC(Q)$:

```{r, echo=FALSE}
D<-10000
A<-500
v<-12.95
r<-0.35

curve(x * v * r / 2 + D * A / x, 1000, 3000,
      ylab=expression(TRC(Q)),
      xlab=expression(Q),
      col='forestgreen')

```

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = D / 52


def f(Q):
  return Q * v * r / 2 + D * A / Q

def grad(Q):
  return v * r / 2 - D * A / Q**2

df = gradient_descent(f, grad, initial_guess, eta=0.1, iter_limit=100000)
  

```

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 4.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

OK, so we converged in `r format(dim(py$df)[1] - 1, big.mark=",")` iterations with an initial guess of $Q_0 = `r py['initial_guess']`$, and a learning rate of $0.10$; with $\eta=0.01$ the algorithm failed to converge in 100,000 iterations! But `r format(dim(py$df)[1] - 1, big.mark=",")` iterations' hardly seems efficient. What happens if we change the initial guess to 1,000?

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = 1000

df = gradient_descent(f, grad, initial_guess, eta=0.1, iter_limit=100000)
  

```


```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 5.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

It *still* requires nearly 30,000 iterations to converge. Let's try a larger learning rate:

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = 1000

df = gradient_descent(f, grad, initial_guess, eta=1.0, iter_limit=100000)
  

```

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 6.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

It's getting better. One more time with $\eta=5.0$:

```{python, echo=FALSE}

D = 10000
A = 500
v = 12.95
r = 0.35

initial_guess = 1000

df = gradient_descent(f, grad, initial_guess, eta=5.0, iter_limit=100000)
  

```

```{r, echo=FALSE}
suppressMessages(require(kableExtra))

caption = 'Example 7.'
n <- dim(py$df)[1]

py$df[c(1:5,(n-5):n),] %>%
  kable(caption = caption, row.names=FALSE, align='c') %>%
    kable_styling(bootstrap_options=c("striped", "hover", "condensed", "responsive"), full_width=FALSE)

```

OK, that's more like it. We converged in only `r format(dim(py$df)[1] - 1, big.mark=",")` iterations.

Just out of curiosity, since we know the exact optimal form for the EOQ model, let's compare:

$$Q^*_{exact} = `r round(sqrt(2 * D * A / (v * r)),3)`$$
```{python, echo=FALSE}

qstar = df.iloc[-1]['$x_n$']

```


$$Q^*_{GD} = `r round(py['qstar'],3)`$$

This example illustrates the challenges associated with using gradient descent, and the sample problem involves only a single decision variable. An additional issue with gradient descent is that it uses a single learning rate for all decision variables.