---
title: "Priority Queues"
author: "Gary Mitchell"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## M/M/c Priority Queues

M/M/c priority queues are simply M/M/c queues with multiple priority classes. So what's a priority class? Think Fastpass vs Standby lines at a Disney park, First Class vs Coach passenger lines at an airline ticket counter. In these examples, there are two classes of service.

Strictly speaking, priority queues don't work exactly like Fastpass/Standby or First Class/Coach lines. Think of a priority queue as a *sorted* queue where all of the priority 1 (the highest priority class) customers are always first in line, followed by all of the priority 2 customers, and so forth. When a new priority 1 customer arrives at the system, s/he advances past all lower priority customers to become the last priority 1 customer in line (i.e., the priority 1 customer most recently arrived at the system) immediately followed by the priority 2 customer that has been in the system the longest. No priority 2 customer will be served until all priority 1 customers in line have been served.

## M/M/c/Priority is still an M/M/c queue 

Because an M/M/c priority queue is still an M/M/c queue, all of the standard metrics that apply to M/M/c queues *still* apply to the M/M/c/Priority queue. When $c > 1$:

$$\lambda_{agg} = \sum_{\lambda_j \in \lambda} \lambda_j$$
$$p_0 = \Biggl(\sum_{n=0}^{c-1}\frac{r^n}{n!} + \frac{r^c}{c!(1-\rho)}\Biggr)^{-1}$$
$$L_q = \frac{r^c \rho}{c!(1-\rho)^2}p_0$$

And, if $c = 1$, we have the simpler forms

$$p_0 = 1 - \rho$$
$$Lq = \frac{\lambda^2}{\mu(\mu - \lambda)}$$

All of this is exactly what we had for standard (i.e., FCFS) M/M/c queues.

## So where do priority queues differ?

M/M/c/priority queues differ only in having *additional* metrics for the priority classes.

$$B_k = 1 - \sum_{j=1}^{k}\frac{\lambda_j}{c \mu}, B_0=1$$

$$W_{q,k} = \frac{(1-\rho)L_q}{\lambda B_{k-1}B_k} = \frac{(1-\rho)W_q}{B_{k-1}B_k}$$
$$L_{q,k} = \lambda_k W_{q,k}$$

## So where's the confusion?##

The M/M/c/FCFS (i.e., the standard M/M/c non-priority queue) was formulated *before* the M/M/c/Priority queue. At the time, the notation from prior queuing models was adopted so that $\lambda$ was defined as the arrival rate. Subsequent to this, the M/M/c/Priority model was developed and had to address having arrival rates (i.e., $\lambda_j$) for each priority class.
 
**Note that, for a priority queue, whenever we use $\lambda$ (as opposed to $\lambda_k$) in an expression, we are treating $\lambda$ as if it's $\lambda_{agg} = \sum_{\lambda_j \in \lambda} \lambda_j$.** 

## An example

Let's say that we have an M/M/c/Priority queue with the following characteristics:

$$\boldsymbol{\lambda} = \{5, 10, 5\}$$
Here, the **bold** $\lambda$ is a notational way of indicated that $\boldsymbol{\lambda}$ is a *set* containing the values 5, 10, and 5 (hence the reason we still reference $\lambda$ in the project specification document, even when we allow it to be multivalued for a priority queue). For pratical reasons, we use a Python *tuple*, instead of a set, because we need indexed access of *ordered* data (i.e., we need to be able to access $\lambda_2$ and have it be the arrival rate for priority class 2). However, in all other respects (other than using a *tuple* instead of a *set*), we are mirroring the M/M/c/FCFS and M/M/c/Priority models in the way we are naming our parameters.

## With numbers

In addition to $\boldsymbol{\lambda} = {5, 10, 5}$, let's assume $\mu = 25$ and $c = 1$. In this case, since we have a single server M/M/c queue (i.e., and M/M/1), we can use the simpler M/M/1 model results to calculate our metrics:

$$L_q = \frac{\lambda^2}{\mu(\mu - \lambda)} = \frac{(\sum_{\lambda_j}\lambda_j)^2}{\mu(\mu - (\sum_{\lambda_j}\lambda_j))} = \frac{\lambda_{agg}^2}{\mu(\mu - \lambda_{agg})} = \frac{20^2}{25(5)} = 3.2\text{ customers}$$
Conceptually, because $L_q$ is the average number of customers waiting *for the entire system*, it makes sense that the arrival rate used here must represent the total arrivals for the system *and not just for a single priority class*. The same idea would apply if we had a multi-server queue, we would simply use the M/M/c formula for $L_q$ instead of the simpler M/M/1 formula.

To find $W_q$ or any other of the major queue metrics, we simply apply Little's Law:

$$L = L_q + r = L_q + \frac{\lambda}{\mu} = 3.2 + 0.8 = 4.0 \text{ customers}$$

$$W_q = \frac{L_q}{\lambda} = \frac{L_q}{\sum_{\lambda_j}\lambda_j} = \frac{3.2}{20} = 0.16 \text{ hours}  = 9.6 \text{ minutes}$$
and

$$W = W_q + \frac{1}{\mu} = 0.16 \text{ hours} + 0.04 \text{ hours} = 0.2 \text{ hours} = 12 \text{ minutes}$$

### Now, by class

Now we can also (want to) calculate queuing metrics by priority class. Little's Law, with some minor tweaks, allows us to do this too:

$$L_{q,k} = \lambda_k W_{q,k}$$

$$L_k = L_{q,k} + \frac{\lambda_k}{\mu} = W_k + \frac{1}{\mu}$$
From these, we can obtain all of the fundamental relationships for the priority classes. So, all we need is $W_{q,k}$ and we're set. Fortunately, $W_{q,k}$ is easily calculated by adjusting $W_q$ for the desired priority class:

$$W_{q,k} = \frac{(1-\rho)W_q}{B_{k-1}B_k}$$
And because we know by Little's Law that $W_q = \frac{L_q}{\lambda}$, then we also have:

$$W_{q,k} =  \frac{(1-\rho)L_q}{\lambda B_{k-1}B_k}$$

Now, because the $Lq$ in the expression is *for the entire* system and not a single service class (i.e., it's not an $L_{q,k}$), then we know that our $\lambda = $lambda_{agg}$. Now, let's say we want to calculate $L_{q,k}$ and $W_{q,k}$ for all three priority classes. We could do so using the following process:

1.    Calculate $B_0, \dots, B_n, n=|\boldsymbol{\lambda}|$
2.    Collect what we already know, i.e., $\lambda_{agg}=20, L_q = 3.2, \rho=0.8$
3.    Using $\lambda_{agg}$ for $\lambda$, calculate $W_{q,k}$.

The table below shows our results:

```{r, echo=FALSE}
suppressPackageStartupMessages(require(kableExtra))

lambda<-c(5, 10, 5)

mu<-25
c<-1

lambda_agg<-sum(lambda)
rho<-lambda_agg / mu
lq<-lambda_agg^2/(mu*(mu-lambda_agg))

rhoj<-cumsum(lambda)/mu

Bk<-1-rhoj
Bkminus1<-c(1, Bk[-3])

wqk<-(1-rho) * lq / (lambda_agg * Bkminus1 * Bk)
lqk<-lambda * wqk

qdat<-data.frame(k=1:length(lambda), Bkminus1=Bkminus1, Bk=Bk, wqk=wqk, lqk=lqk)
colnames(qdat)<-c("$k$", "$B_{k-1}$","$B_k$", "$W_{q,k}$", "$L_{q,k}$")

qdat %>%
  kbl() %>%
  kable_styling(full_width=FALSE)

```

Now we can even relate our results by priority class back to the overall queue metrics. Because the $L_{q,k}$ represent the average number of customers in priority class $k$, we can calculate:

$$L_q = \sum_{k = 1}^{|\boldsymbol{\lambda}|}L_{q,k}$$
Additionally, because $W_{q,k}$ is the average time spent waiting for a customer in priority class $k$, we can calculate the overall average waiting time, $W_q$, as the weighted average of the $W_{q,k}$:

$$W_q = \frac{\sum_{k=1}^{|\boldsymbol{\lambda}|}\lambda_k W_{q,k}}{\sum_{k=1}^{|\boldsymbol{\lambda}|} \lambda_k}$$
or alternatively:

$$W_q = \sum_{k=1}^{|\boldsymbol{\lambda}|} p_k W_{q,k}$$
where $p_k = \frac{\lambda_k}{\sum_{k=1}^{|\boldsymbol{\lambda}|}\lambda_k}$. In other words, $W_q$ is the *arrival rate weighted average* waiting time.

Using these formulas, we calculate $W_{q,k} = `r lambda %*% wqk / lambda_agg`$ and $L_{q,k} = `r sum(lqk)`$.

It's comforting to see that our overall queue metrics calculated in this way match the values we calculated directly using the *overall* formuas earlier. However, it's more important than that. Calculating $Lq$ as the sum of the $L_{q,k}$ and $W_q$ as the weighted average of the $W_{q,k}$ emphasizes the nature of these metrics, their relationship to each other, and the values of $\lambda$ appropriate for each of the calculations.

## Projects 3 and 4

The specification for Projects 3 and 4 are very clear in their treatment of $\lambda$ values:

*   $\lambda$ always represents the *queue*-level arrival rate(s). If the queuing discipline is FCFS (i.e., *not* a priority queue), then $\lambda$ is a single value. If the queuing discipline is *priority*, then $\lambda$ is a tuple, representing the set of arrival rates $\boldsymbol{\lambda}$.
*   $\lambda_k$ (or $\lambda_j$ or $\lambda_i$) always represents the arrival rate for a specific service class designated by the subscript (e.g., $\lambda_2$ is the priority class 2 arrival rate).

Clearly, if $\lambda$ is multivalued, you must aggregate the $\lambda$ values for any queue-level calculation (e.g., $L_q$, $\rho$, etc.). When a call signature calls for $\lambda$, it's asking for the whole thing (including the tuple if $\lambda$ is multivalued). If the spec shows $\lambda_k$, then it's calling for $\lambda_k$, i.e., the class $k$ arrival rate.

Have a look at the Project 4 function call signatures (the function bodies have been omitted here because... well because that's the work for Project 4).

```{python, exec=FALSE}
def calc_bk_mmc(k, lamda, mu, c=1):
  pass

def calc_wqk_mmc(k, lamda, mu, c=1):
  pass

def calc_lqk_mmc(k, lamda, wqk):
  pass

def use_littles_law(lamda, mu, c=1, **kwargs):
  pass
```

*lamda* clearly represents the tuple of $\lambda$ values. How do you know? There is no $k$ subscript (i.e., it is referencing lamda, not lamda_k). If a calculation *inside* the function needs $\lambda_k$, then it simply obtains it from the tuple; if you have the tuple, you can access *any* of the individual priority class $\lambda$'s, right?

## Projects 5 and 6

The handling of $\lambda$ will be a little different our OOP implementations. In these, only priority queues need $\lambda_k$; all queues except the priority queue only need a single $\lambda$ representing the single arrival rate for the queue. The M/M/1 and M/M/c queue classes will inherit this attribute from their base class (you'll understand what this means after we discuss object-oriented). If the value of $\lambda$ provided is multi-valued, then all but the priority queue class can simply sum the values and store the aggregate as $\lambda$. The priority queue class will *additionally store the multi-valued $\lambda$ as $\lambda_k$.

All of this should make more sense to you after you read the Project 5 & 6 description and we've covered object-oriented programming in class. In the meantime, just remember this discussion is here when you're working on projects 5 & 6. 

## Summary

OK, that about wraps it up. Hopefully this makes things more clear, but here's the condensed version:

*   If the spec asks for lamda in the call signature, it is lamda for the entire queue, not a single priority class
*   If the queue is a *priority* queue, then that lamda requested is the tuple
*   If a function needs $\lambda_k$, it will index into lamda using $k$ to get the $\lambda_k$ it needs.











