---
title: "Probability Cheat Sheet"
author: "Gary Mitchell"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This cheat sheet provides a handy reference to the probability topics covered in Decision Modeling & Analytics and which will be useful in future courses such as Project Management, Supply Chain Analytics, Process Analysis, Design, and Quality Management, Foundations of Operations Management, etc.


## Probability Notation

The following notation will be used throughout.


```{r, echo=FALSE}
library(readxl)
library(kableExtra)

tab<-read_excel("Probability Notation.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[, -1]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

## Laws of Probability

### 1st Law of Probability

-	The probability of any event is a number between zero and one: $0 \leq F(x) \leq 1$
-	A larger probability corresponds to the intuitive notion of greater likelihood.
-	An event whose associated probability is 1.0 is virtually certain.
-	An event whose probability is 0.0 is virtually certain not to occur.

The first law gives us the concept of complementary probability:

- Discrete Distributions: $Pr\{X > x\} = \bar{F}(x) = 1 - F(x) = 1 - Pr\{X \leq x\}$
- Continuous Distributions: $Pr\{X \geq x\} = \bar{F}(x) = 1 - F(x) = 1 - Pr\{X \leq x\}$

### 2nd Law of Probability

If $A$ and $B$ are *mutually exclusive* events, then 
$$Pr\{A \cup B\} = Pr\{A \thinspace or \thinspace B\}= Pr\{A\} + Pr\{B\}$$. 

Note: when $A$ and $B$ are *not* mutually exclusive, then $Pr\{A \cup B\} = Pr\{A\} + Pr\{B\} - Pr\{A \cap B\}$

### 3rd Law of Probability

The probability of one event *given* that another event has occurred is called a conditional
probability. Given events $A$ and $B$, $Pr\{A|B\}$ is read as *the probability of $A$ given $B$*.

$$Pr\{A|B\} = \frac{Pr\{A \cup B\}}{Pr\{B\}}$$
from which it follows that:

$$Pr\{A \cup B\} = Pr\{A|B\}Pr\{B\}$$

#### The Complement of a Conditional Probability

Given an event $A$ and its complement, $A^C$m we know that $Pr\{A^C\} = 1 - Pr\{A\}$.
But what does it mean to calculate the complement of a conditional probability?

$$Pr\{A^C|B\} = 1 - Pr\{A|B\}$$
Note that the complementary conditional probability is always calculatd *within*
the conditional event, i.e. the conditional event remains the same for both
the conditional probability and its complement.

### 4th Law of Probability

For two *independent* events:

$$Pr\{A \cap B\} = Pr\{A\}Pr\{B\}$$
This follows from the definition of the $3^{rd}$ Law. If two events are independent,
then the occurrence of one does not affect the probability of the other, i.e.
$Pr\{A|B\} = Pr\{A\}$.

This forms the basis for a classic test of independence.

>**If** $Pr\{A \cap B\} = Pr\{A\}Pr\{B\}$ **then** $A$ and $B$ are *independent*.

### Law of Total Probability

The law of total probability follows from the $1^{st}$ and $2^{nd}$ laws. Given
mutually exclusive events $A_1$, $A_2$, $\dots$, $A_n$ so that 
$Pr\{A\} = Pr\{A_1\} + Pr\{A_1\} + \dots + Pr\{A_n\}$, then:

$$Pr\{B\} = Pr\{B|A_1\}Pr\{A_1\} + Pr\{B|A_2\} Pr\{A_2\} + \dots + Pr\{B|A_n\} Pr\{A_n\}$$
Using the $2^{nd}$ Law, we see that this is equivalent to:

$$Pr\{B\} = Pr\{B \cup A_1\} + Pr\{B \cup A_2\} + \dots + Pr\{B \cup A_n\}$$

### Bayes' Theorem

Bayes' Theorem is a logical extension of the definition of conditional probability using the Law of Total Probability. From the $3^{rd}$ Law, we know that:

$$Pr\{A_1|B\} = \frac{Pr\{A_1 \cup B\}}{Pr\{B\}}$$
Also, by the $3^{rd}$ Law, we can substitute $Pr\{B|A\}Pr\{A\}$ for $Pr\{A \cup B\}$:

$$Pr\{A_1|B\} = \frac{Pr\{B|A_1\}Pr\{A_1\}}{Pr\{B\}}$$
This is a complete representation of Bayes' Theorem. However, many statistics texts 
present a form that uses the Law of Total Probability to calculate the denominator,
$Pr\{B\}$:

$$Pr\{B\} = Pr\{B|A_1\}Pr\{A_1\} + Pr\{B|A_2\} Pr\{A_2\} + \dots + Pr\{B|A_n\} Pr\{A_n\}$$
and substituting this we obtain:

$$Pr\{A_1|B\} = \frac{Pr\{B|A_1\}Pr\{A_1\}}{Pr\{B|A_1\}Pr\{A_1\} + Pr\{B|A_2\} Pr\{A_2\} + \dots + Pr\{B|A_n\} Pr\{A_n\}}$$
This looks intimidating, but it's really just our original definition for conditional probability, with some substitutions.


## Discrete Distributions

For a discrete distribution, the set of outcomes in $\Omega$ is finite and countable.
For this reason, $Pr\{X = x\} = m(x) \geq 0$ (whereas $Pr\{X = x\} \equiv 0$ for
continuous distributions).

### Discrete Uniform

#### Properties

- Set of possible outcomes $\Omega$
- $m(x) = $\frac{1}{|\Omega|}, for every $x \in \Omega$

where $n = |\Omega|$, i.e. the cardinality of $\Omega$
(note: for a set $S$, $|S|$, the cardinality of $S$, is the number of elements in $S$).

#### Calculations

```{r, echo=FALSE}
dist<-"Discrete Uniform"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

### Poisson

#### Properties

- Non-negative integer distribution
- Single parameter $\lambda$, the *rate*
- Markovian (memoryless) property

$\lambda$ must be a *rate*, meaning it has units of measure quantity per interval
(e.g. customers per hour).

#### Calculations

```{r, echo=FALSE}
dist<-"Poisson"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

### Binomial

#### Properties

- Binary outcome space (e.g. Defective/Not Defective, True/False, etc.)
- $n$ independent trials
- Event of interest (often referred to as a success) occurs with probability $p$
- Models probability of $x$ occurances in $n$ trials


#### Calculations

```{r, echo=FALSE}
dist<-"Binomial"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

## Continuous Distributions

For a continuous distribution, the set of outcomes in $\Omega$ is infinite.
$Pr\{X = x\} \equiv 0$ for all continuous distributions because:

$$Pr\{X = x\} = \int_{x}^{x}f(t)dt = \bigl| F(x) \bigr|_x^x = F(x) - F(x) = 0$$

### Continuous Uniform

#### Properties

- Bounded above and below
- Symmetrical
- Every outcome occurs with equal likelihood
- Two parameters
  - Minimum (lower bound) a
  - Maximum (upper bound) b
- $Pr\{X < a\} = 0$
- $Pr\{X > b\} = 0$


#### Calculations

```{r, echo=FALSE}
dist<-"Continuous Uniform"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

### Normal (Gaussian)

#### Properties

- Unbounded
- Symmetrical
- Two parameters
  - mean $\mu$
  - standard deviation $\sigma
  
Note: no closed form expressions for $F(x)$ and $F^{-1}(x)$.


#### Calculations

```{r, echo=FALSE}
dist<-"Normal"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

### Exponential

#### Properties

- Non-negative distribution
- Bounded below at $0$
- Unbounded above
- Asymmetrical
- Single parameter $\lambda$, the *rate*
- Markovian (memoryless) property.


#### Calculations

```{r, echo=FALSE}
dist<-"Exponential"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```

### Shifted exponential


#### Properties

- Unbounded above
- Bounded below by *shift*, $a$
- Asymmetrical
- Single parameter $\lambda$, the *rate*.
  

#### Calculations

```{r, echo=FALSE}
dist<-"Shifted Exponential"

tab<-read_excel("Probability Calculations.xlsx")
tab[is.na(tab)]<-""
ptab<-tab[tab$Distribution == dist, 2:4]
ptab %>%
  kable(booktabs=TRUE) %>%
  kable_styling("striped")
```


## Generating Random Numbers

Random Numbers can be easily generated for any distribution known to R using r*dist*($n$, $\cdot$). where *dist* is the abbreviation for the distribution (e.g. norm for normal, pois for Poisson, etc.), and $n$ is the number of random variates you wish to generate.

For example, we can generate 10 Poisson random variates with mean 5 using:

```{r}
rpois(10, 5)
```

Similarly, we can generate 10 shifted exponential random variates with shift 2 and mean random portion of 1.5 using:

```{r}
2 + rexp(10, 1/1.5)
```



### Inverse Transform Method

Alternatively, we can use the inverse transform method for generating random numbers. Given a random variable $X$ with inverse cumulative probability function $F^{-1}(\alpha)$, we can generate a random variate $x$ using:

$$x = F^{-1}(U)$$

where $U \sim Unif[0,1]$, i.e. $U$ is a uniformly distributed random number between 0 and 1.

For example, given $U = \{0.7294, 0.4868, 0.3845, 0.0068, 0.0037, 0.9949, 0.1079, 0.4187, 0.7179, 0.7425\}$ we would generate shifted exponential random variates with shift 2 and mean 3.5 as follows:

```{r, echo=FALSE}
U<-c(0.7294, 0.4868, 0.3845, 0.0068, 0.0037, 0.9949, 0.1079, 0.4187, 0.7179, 0.7425)
```

```{r}
a<-2
mu<-3.5
lambda<-1/(mu-a)

a + qexp(U, lambda)
```

## Covariance and Correlation

For two jointly distributed random variables $X$ and $Y$, we calculate the covariance and correlation to gain insight in how the RVs vary relative to each other.

### Discrete Random Variables

$$COV(X, Y) = \sum_{(x,y) \in \Omega}Pr\{X=x, Y=y\}(x - E[X])(y - E[Y])$$
Alternatively, we can write:

$$COV(X, Y) = \sum_{(x,y) \in \Omega}m(x, y)(x - E[X])(y - E[Y])$$
and the correlation is given by:

$$CORR(X,Y) = \frac{COV(X,Y)}{S[X]S[Y]}$$

### Continuous Random Variables

$$COV(X, Y) = \int \int (x - E[X])(y - E[Y])f(x, y)\thinspace dx\thinspace dy$$
$CORR(X,Y)$ is calculated as shown above for discrete distributions.



